Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-13 ind_files/99_Appendix_A.tex
Change notation from dB to dW; Change population vector notation from v(t) to n(t)

Consider a one-dimensional real It\^{o} process given by $dX_t = \mu(X_t,t)dt + \sigma(X_t,t)dB_t$ on a filtered probability space $\Omega \subseteq \mathbb{R}$ that admits a probability density function $P(x,t)$ on $\Omega \times [0,\infty)$. Let $f:\mathbb{R}\to\mathbb{R}$ be an arbitrary $C^2(\mathbb{R})$ function. By It\^{o}'s formula, we have:

@@

    df(X_t) = \left(\mu f' + \frac{\sigma^2}{2}f''\right)dt + \sigma f' dB_t

@@

    \mathbb{E}[f(X_t)] = \mathbb{E}\left[\int\limits_{0}^{t}\left(\mu f' + \frac{\sigma^2}{2}f''\right)ds\right] + \mathbb{E}\left[\int\limits_{0}^{t}\sigma f' dB_s\right]

@@

Equation \eqref{FPE} is the Fokker-Planck equation in one dimension. Using the exact same strategy, the multidimensional Fokker-Planck equation for the $m$-dimensional It\^{o} Process $d\mathbf{X}_t = \boldsymbol{\mu}(\mathbf{X}_t,t)dt + \boldsymbol{\sigma}(\mathbf{X}_t,t)d\mathbf{B}_t$ can be easily found to be:

>>>>

Consider a one-dimensional real It\^{o} process given by $dX_t = \mu(X_t,t)dt + \sigma(X_t,t)dW_t$ on a filtered probability space $\Omega \subseteq \mathbb{R}$ that admits a probability density function $P(x,t)$ on $\Omega \times [0,\infty)$. Let $f:\mathbb{R}\to\mathbb{R}$ be an arbitrary $C^2(\mathbb{R})$ function. By It\^{o}'s formula, we have:

@@

    df(X_t) = \left(\mu f' + \frac{\sigma^2}{2}f''\right)dt + \sigma f' dW_t

@@

    \mathbb{E}[f(X_t)] = \mathbb{E}\left[\int\limits_{0}^{t}\left(\mu f' + \frac{\sigma^2}{2}f''\right)ds\right] + \mathbb{E}\left[\int\limits_{0}^{t}\sigma f' dW_s\right]

@@

Equation \eqref{FPE} is the Fokker-Planck equation in one dimension. Using the exact same strategy, the multidimensional Fokker-Planck equation for the $m$-dimensional It\^{o} Process $d\mathbf{X}_t = \boldsymbol{\mu}(\mathbf{X}_t,t)dt + \boldsymbol{\sigma}(\mathbf{X}_t,t)d\mathbf{W}_t$ can be easily found to be:

### 2023-03-13 ind_files/99_Appendix_B.tex
Change notation from dB to dW; Change population vector notation from v(t) to n(t)

d\mathbf{X}_t = \boldsymbol{\mu}(\mathbf{X}_t)dt + \boldsymbol{\sigma}(\mathbf{X}_t)d\mathbf{B}_t

@@

df(\mathbf{X}_t) = \left[\left(\nabla_{\mathbf{X}}f\right)^{\mathrm{T}}\boldsymbol{\mu} + \frac{1}{2}\mathrm{Tr}[\boldsymbol{\sigma}^{\mathrm{T}}(H_{\mathbf{X}}f)\boldsymbol{\sigma}]\right]dt + \left(\nabla_{\mathbf{X}}f\right)^{\mathrm{T}}\boldsymbol{\sigma}d\mathbf{B}_t

@@

Finally, denoting $d\mathbf{B}_t = [dB^{(1)}_t,dB^{(2)}_t, \ldots, dB^{(m)}_t]^{\mathrm{T}}$ where each $dB^{(j)}_t$ is an independent one dimensional Wiener process, we have:

@@

\left(\mathbf{D}d\mathbf{B}_t\right)_j &= \sum\limits_{k=1}^{m}\mathbf{D}_{jk}dB^{(k)}_t\nonumber\\
&= \sum\limits_{k=1}^{m}\delta_{jk}\left(A^{+}_{j}A^{+}_{k}\right)^{\frac{1}{4}}dB^{(k)}_t\\
&= \left(A^{+}_{j}\right)^{1/2}dB^{(j)}_t

@@

\frac{1}{\sqrt{K}}\left(\nabla_{\mathbf{X}}f\right)^{\mathrm{T}}\mathbf{D}d\mathbf{B}_t &= \frac{1}{\sqrt{K}}\sum\limits_{j=1}^{m}\left(\nabla_{\mathbf{x}} f_i\right)_j\left(\mathbf{D}d\mathbf{B}_t\right)_j\nonumber\\
&=  \frac{1}{N\sqrt{K}}\sum\limits_{j=1}^{m}\left(\delta_{ij}-p_i\right)\left(A^{+}_{j}\right)^{1/2}dB^{(j)}_t\\
&= \frac{1}{N\sqrt{K}}\left(A^{+}_{i}\right)^{1/2}dB^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}\right)^{1/2}dB^{(j)}_t\label{nD_for_Ito_third_term}

@@

&+ \frac{1}{\sqrt{K} N(t)}\underbrace{\left[\left(A^{+}_{i}\right)^{1/2}dB^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}\right)^{1/2}dB^{(j)}_t\right]}_{\substack{\text{Non-directional finite size effects}\\\text{due to stochastic fluctuations}}}

>>>>

d\mathbf{X}_t = \boldsymbol{\mu}(\mathbf{X}_t)dt + \boldsymbol{\sigma}(\mathbf{X}_t)d\mathbf{W}_t

@@

df(\mathbf{X}_t) = \left[\left(\nabla_{\mathbf{X}}f\right)^{\mathrm{T}}\boldsymbol{\mu} + \frac{1}{2}\mathrm{Tr}[\boldsymbol{\sigma}^{\mathrm{T}}(H_{\mathbf{X}}f)\boldsymbol{\sigma}]\right]dt + \left(\nabla_{\mathbf{X}}f\right)^{\mathrm{T}}\boldsymbol{\sigma}d\mathbf{W}_t

@@

Finally, denoting $d\mathbf{W}_t = [dW^{(1)}_t,dW^{(2)}_t, \ldots, dW^{(m)}_t]^{\mathrm{T}}$ where each $dW^{(j)}_t$ is an independent one dimensional Wiener process, we have:

@@

\left(\mathbf{D}d\mathbf{W}_t\right)_j &= \sum\limits_{k=1}^{m}\mathbf{D}_{jk}dW^{(k)}_t\nonumber\\
&= \sum\limits_{k=1}^{m}\delta_{jk}\left(A^{+}_{j}A^{+}_{k}\right)^{\frac{1}{4}}dW^{(k)}_t\\
&= \left(A^{+}_{j}\right)^{1/2}dW^{(j)}_t

@@

\frac{1}{\sqrt{K}}\left(\nabla_{\mathbf{X}}f\right)^{\mathrm{T}}\mathbf{D}d\mathbf{W}_t &= \frac{1}{\sqrt{K}}\sum\limits_{j=1}^{m}\left(\nabla_{\mathbf{x}} f_i\right)_j\left(\mathbf{D}d\mathbf{W}_t\right)_j\nonumber\\
&=  \frac{1}{N\sqrt{K}}\sum\limits_{j=1}^{m}\left(\delta_{ij}-p_i\right)\left(A^{+}_{j}\right)^{1/2}dW^{(j)}_t\\
&= \frac{1}{N\sqrt{K}}\left(A^{+}_{i}\right)^{1/2}dW^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}\right)^{1/2}dW^{(j)}_t\label{nD_for_Ito_third_term}

@@

&+ \frac{1}{\sqrt{K} N(t)}\underbrace{\left[\left(A^{+}_{i}\right)^{1/2}dW^{(i)}_t - p_i\sum\limits_{j=1}^{m}\left(A^{+}_{j}\right)^{1/2}dW^{(j)}_t\right]}_{\substack{\text{Non-directional finite size effects}\\\text{due to stochastic fluctuations}}}

### 2023-03-13 ind_files/99_Appendix_C.tex
Minor phrasing and notation changes
Change notation from dB to dW; Change population vector notation from v(t) to n(t)

&+\frac{1}{\sqrt{K}N_{K}(t)}\left(\sum\limits_{i=1}^{m}\left(f^2_i - \overline{f^2}\right)\sqrt{A_i^+}dB_{t}^{(i)}\right)

@@

For the third term, we need to use It\^{o}'s formula. Here, the relevant version of It\^{o}'s formula is the one-dimensional version of \eqref{nD_Ito_formula}. Given a one-dimensional process $dX_t = S(X_t)dt + \sum D_j(X_t)dB^{(j)}_t$ with $S, D_j$ being suitable real functions and each $B^{(j)}_t$ being an independent Wiener process, It\^{o}'s formula says that given any $C^2(\mathbb{R})$ function $g(x)$, we have the relation:

@@

dg(X_t) = \left(S(X_t)g'(X_t) + \frac{g''(X_t)}{2}\sum\limits_{j}D_j^2(X_t)\right)dt + \sum\limits_{j}D_j(X_t)g'(X_t)dB^{(j)}_t 

@@

d(\overline{f}^2) = \left(2\overline{f}S(X_t) + \sum\limits_{j}D_j^2(X_t)\right)dt + \sum\limits_{j}2\overline{f}D_j(X_t)dB^{(j)}_t

@@

where the relevant functions $S$ and $D_j$ can be read off from \eqref{nD_stochastic_Price}. Since the $dB$ terms are unwieldy, we will denote the contribution of all the $dB_t$ terms collectively by $dB_{\sigma^2_{f}}$ to reduce notational clutter and only explicitly calculate these terms at the end. We also note that the covariance operator is a bilinear form, \emph{i.e.} given any three quantities $X$, $Y$ and $Z$ and any constant $a \neq 0$, we have the relations:

@@

&+ dB_{\sigma^2_{f}}

@@

&+ 2\textrm{Cov}\left(\frac{\partial f}{\partial t},f\right)dt + M_{\sigma^2_f}(\mathbf{p},N_K)dt + dB_{\sigma^2_{f}}

@@

& + 2\textrm{Cov}\left(\frac{\partial f}{\partial t},f\right)dt + M_{\sigma^2_f}(\mathbf{p},N_K)dt + dB_{\sigma^2_{f}}

@@

\label{app_stochastic_Price_variance}

@@

This is precisely equation \eqref{stochastic_Price_variance} in the main text. To calculate the mutation term, we substitute \eqref{intermediate_2_for_variances} into \eqref{intermediate_1_for_variances} to find

@@

which is equation \eqref{variance_price_mutation_term} in the main text. For the $dB$ terms, we can use equations \eqref{2nd_term_for_variances} and \eqref{3rd_term_for_variances} to calculate:

@@

dB_{\sigma^2_{f}} &=  \frac{1}{\sqrt{K}N_{K}(t)}\left(\sum\limits_{i=1}^{m}\left(f^2_i - \overline{f^2} - 2 \overline{f}(f_i - \overline{f})\right)\sqrt{A_i^+}dB_{t}^{(i)}\right)\\
&= \frac{1}{\sqrt{K}N_{K}(t)}\left(\sum\limits_{i=1}^{m}\left(f^2_i - \overline{f^2} - 2 \overline{f}f_i - 2\overline{f}^2\right)\sqrt{A_i^+}dB_{t}^{(i)}\right)\\
&= \frac{1}{\sqrt{K}N_{K}(t)}\left(\sum\limits_{i=1}^{m}\left(f_i - \overline{f}\right)^2\sqrt{A_i^+}dB_{t}^{(i)}\right)

>>>>

&+\frac{1}{\sqrt{K}N_{K}(t)}\left(\sum\limits_{i=1}^{m}\left(f^2_i - \overline{f^2}\right)\sqrt{A_i^+}dW_{t}^{(i)}\right)

@@

For the third term, we need to use It\^{o}'s formula. Here, the relevant version of It\^{o}'s formula is the one-dimensional version of \eqref{nD_Ito_formula}. Given a one-dimensional process $dX_t = S(X_t)dt + \sum D_j(X_t)dW^{(j)}_t$ with $S, D_j$ being suitable real functions and each $W^{(j)}_t$ being an independent Wiener process, It\^{o}'s formula says that given any $C^2(\mathbb{R})$ function $g(x)$, we have the relation:

@@

dg(X_t) = \left(S(X_t)g'(X_t) + \frac{g''(X_t)}{2}\sum\limits_{j}D_j^2(X_t)\right)dt + \sum\limits_{j}D_j(X_t)g'(X_t)dW^{(j)}_t 

@@

d(\overline{f}^2) = \left(2\overline{f}S(X_t) + \sum\limits_{j}D_j^2(X_t)\right)dt + \sum\limits_{j}2\overline{f}D_j(X_t)dW^{(j)}_t

@@

where the relevant functions $S$ and $D_j$ can be read off from \eqref{nD_stochastic_Price}. Since the $dW$ terms are unwieldy, we will denote the contribution of all the $dW_t$ terms collectively by $dW_{\sigma^2_{f}}$ to reduce notational clutter and only explicitly calculate these terms at the end. We also note that the covariance operator is a bilinear form, \emph{i.e.} given any three quantities $X$, $Y$ and $Z$ and any constant $a \neq 0$, we have the relations:

@@

&+ dW_{\sigma^2_{f}}

@@

&+ 2\textrm{Cov}\left(\frac{\partial f}{\partial t},f\right)dt + M_{\sigma^2_f}(\mathbf{p},N_K)dt + dW_{\sigma^2_{f}}

@@

& + 2\textrm{Cov}\left(\frac{\partial f}{\partial t},f\right)dt + M_{\sigma^2_f}(\mathbf{p},N_K)dt + dW_{\sigma^2_{f}}

@@

\label{app_nD_stochastic_Price_variance}

@@

This is precisely equation \eqref{nD_stochastic_Price_variance} in the main text. To calculate the mutation term, we substitute \eqref{intermediate_2_for_variances} into \eqref{intermediate_1_for_variances} to find

@@

which is equation \eqref{variance_price_mutation_term} in the main text. For the $dW$ terms, we can use equations \eqref{2nd_term_for_variances} and \eqref{3rd_term_for_variances} to calculate:

@@

dW_{\sigma^2_{f}} &=  \frac{1}{\sqrt{K}N_{K}(t)}\left(\sum\limits_{i=1}^{m}\left(f^2_i - \overline{f^2} - 2 \overline{f}(f_i - \overline{f})\right)\sqrt{A_i^+}dW_{t}^{(i)}\right)\\
&= \frac{1}{\sqrt{K}N_{K}(t)}\left(\sum\limits_{i=1}^{m}\left(f^2_i - \overline{f^2} - 2 \overline{f}f_i - 2\overline{f}^2\right)\sqrt{A_i^+}dW_{t}^{(i)}\right)\\
&= \frac{1}{\sqrt{K}N_{K}(t)}\left(\sum\limits_{i=1}^{m}\left(f_i - \overline{f}\right)^2\sqrt{A_i^+}dW_{t}^{(i)}\right)

### 2023-03-13 ind_files/99_Appendix_D_examples.tex
Change notation from dB to dW; Change population vector notation from v(t) to n(t)

\section{An example in one dimension: The stochastic logistic equation}
Here, we analyze example \ref{ex_1D_stoch_logistic}. To recap, we had a population of individuals that exhibit a constant per-capita birth rate $\lambda > 0$, and a per-capita death rate that had the linear density-dependence $d(n) = \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n$, where $\mu$ and $K$ are positive constants. Thus, we have the equations
\begin{equation}
\label{ex_1D_stoch_logistic_BD_eqns}
\begin{aligned}
    b(n) &= \lambda n\\
    d(n) &= \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n
\end{aligned}
\end{equation}
Here, $K$ is the system-size parameter. Introducing the new variable $x=n/K$, we obtain
\begin{align*}
    b_K(x) &= \frac{1}{K}b(n) = \frac{1}{K}\lambda Kx\\
    d_K(x) &= \frac{1}{K}d(n) = \frac{1}{K}\left(\mu + (\lambda-\mu)\frac{Kx}{K}\right)Kx
\end{align*}
Thus, we have
\begin{equation*}
    A^{\pm}(x) = b_K(x)\pm d_K(x) = x\left(\lambda \pm \left(\left(\mu + (\lambda-\mu)x\right)\right) \right)
\end{equation*}
Defining $r=\lambda-\mu$ and $v=\lambda+\mu$ and using equation \eqref{1D_SDE}, we 
see that the `mesoscopic view' of the system is given by the solution of the SDE
\begin{equation}\label{ex_1D_stoch_logistic_full_SDE}
dX_t =  rX_t(1-X_t)dt + \sqrt{\frac{X_t(v+rX_t)}{K}}dB_t
\end{equation}
From equation \eqref{1D_det_limit}, we see that the deterministic dynamics are
\begin{equation}\label{ex_1D_stoch_logistic_det_limit}
\frac{dx}{dt} = A^-(x) = rx(1-x)
\end{equation}
showing that in the infinite population limit, we obtain the logistic equation. This derivation also makes it clear that two systems with very different stochastic dynamics can nevertheless converge to the same infinite population limit \eqref{ex_1D_stoch_logistic_det_limit}, since \eqref{ex_1D_stoch_logistic_det_limit}t only depends on the difference $\lambda-\mu$.
\myfig{0.6}{Media/App_stoch_logistic_same_w_different_t.png}{Comparison of time series (Gillespie simulations) for two different parameter values of the stochastic logistic equation. Simulations with the parameter values $\lambda = 4, \mu = 2$ are plotted in red, and simulations with the parameter values $\lambda = 31, \mu = 29$ are plotted in blue. Both simulations have $K=200$. The infinite population limit of both simulations is plotted in black dotted lines. The graph shows 10 realizations each for the two sets of parameter values.}{fig_1D_same_w_different_t_comparison}
To illustrate the effects of this seemingly innocent fact, figure \ref{fig_1D_same_w_different_t_comparison} compares two simulations which have the same value of $\lambda - \mu$ but a ten-fold difference in $\lambda+\mu$. As is clear from the figure, though both populations have the same behavior at the infinite population limit, populations with a higher value of $\lambda+\mu$ exhibit much wilder fluctuations and are therefore more prone to stochastic extinction. This is the root cause of the noise-induced selection that occurs in higher dimensions, discussed in detail in part \ref{part_summary}.

@@

    dY_t &= A^-_1(t)Y_tdt + \sqrt{A^+_0(t)}dB_t\nonumber\\
    \Rightarrow dY_t &= r(1 - 2\alpha(t))Y_tdt + \sqrt{\alpha(t)(v+r\alpha(t))}dB_t\label{ex_1D_stoch_logistic_WNA}

@@

Let us imagine a population with $m$ types of individuals that are interacting according to some ecological rules. Let the state of the population be characterized by the vector $\mathbf{v}(t) = [v_1(t),v_2(t),\ldots,v_m(t)]$, where $v_i(t)$ is the number of type $i$ individuals at time $t$. Let the birth and death rates of the $i$th type be given by:
\begin{equation}
\label{nD_example_numbers_b_d_rates}
\begin{aligned}
b_i(\mathbf{v}) &= \left(\lambda + \frac{1}{K}\left(\sum\limits_{j=1}^{m}\beta_{ij}v_j\right)\right)v_i\\
d_i(\mathbf{v}) &= \left(\mu + \frac{1}{K}\left(\sum\limits_{j=1}^{m}\delta_{ij}v_j\right)\right)v_i
\end{aligned}
\end{equation}
where $K > 0$ is our system size parameter (and represents a global carrying capacity across all types), $\lambda > 0$ and $\mu > 0$ are suitable positive constants representing the baseline natality and mortality common to all types, and $\beta_{ij}$ and $\delta_{ij}$ are constants describing the effect of type $j$ individuals on the birth and death rate of type $i$ individuals respectively. The sign of $M_{ij} \coloneqq \beta_{ij} - \delta_{ij}$ determines whether type $j$ has a net positive or negative effect on the growth of type $i$. In ecological communities, this is a per-capita ecological interaction effect. In game-theoretic terms, we can interpret $M_{ij}$ as the payoff obtained by a type $j$ individual playing against a type $i$ individual. I assume that $| M_{ij} | \ll K$. The values $M_{ij}$ are often collected in an $m \times m$ matrix $\mathbf{M}$ called the `payoff matrix' (in evolutionary game theory) or `interaction matrix' (in Lotka-Volterra models). Lotka-Volterra models also frequently assume that the diagonal elements $M_{ii}$ are all equal, though I will not make that assumption here.

Going from population numbers $\mathbf{v}$ to densities $\mathbf{x} = \mathbf{v}/K$, we obtain the birth and death rates:
\begin{equation}
\label{nD_example_density_b_d_rates}
\begin{aligned}
b^{(K)}_i(\mathbf{x}) &= \left(\lambda + \sum\limits_{j=1}^{m}\beta_{ij}x_j\right)x_i\\
d^{(K)}_i(\mathbf{x}) &= \left(\mu + \sum\limits_{j=1}^{m}\delta_{ij}x_j\right)x_i
\end{aligned}
\end{equation}
Thus, we have
\begin{equation*}
A^{\pm}_{i} = x_i\left(\left(\lambda \pm \mu\right) + \sum\limits_{j=1}^{m}\left(\beta_{ij} \pm \delta_{ij}\right)x_j\right)
\end{equation*}
Defining $r \coloneqq \lambda - \mu$, $\nu \coloneqq \lambda + \mu$, and $T_{ij} \coloneqq \beta_{ij} + \delta_{ij}$, we see from equation \eqref{nD_Ito_SDE} that the mesoscopic view is the $m$ dimensional SDE given by
\begin{equation}
\label{nD_example_SDE}
d\mathbf{X}_{t} = \mathbf{A^-}(\mathbf{X}_t)dt + \frac{1}{\sqrt{K}}\mathbf{D}(\mathbf{X}_t)d\mathbf{B}_t
\end{equation}
where 
\begin{equation*}
\mathbf{A^-}_i = {(\mathbf{X}_{t})}_i(r + \sum\limits_{j=1}^{m}M_{ij}{(\mathbf{X}_{t})}_j) 
\end{equation*}
and
\begin{equation*}
(\mathbf{D}\mathbf{D}^{\mathrm{T}})_i = {(\mathbf{X}_{t})}_i(\nu + \sum\limits_{j=1}^{m}T_{ij}{(\mathbf{X}_{t})}_j) 
\end{equation*}
From \eqref{nD_det_limit}, we see that the deterministic limit is a set of $m$ coupled ODEs given by
\begin{equation}
\label{nD_example_det_limit}
\frac{d x_i}{dt} = x_i\left(r + \sum\limits_{j=1}^{m}M_{ij}x_j\right)
\end{equation}
These are precisely the Lotka-Volterra equations for a system of $m$ species. By matching the terms of \eqref{nD_example_density_b_d_rates} with those of \eqref{nD_functional_forms_for_replicator}, we can identify that we have $\mu = 0$ and
\begin{equation}
\label{nD_example_fitness}
w_i(\mathbf{x}) = r + \sum\limits_{j=1}^{m}M_{ij}x_j
\end{equation}
If $p_i(t)$ is the frequency of type $i$ individuals in the population at time $t$ and $N_K(t) = \sum_i x_i(t)$, then the mean fitness is given by
\begin{align}
\overline{w}(t) &= \sum\limits_{i=1}^{m}w_ip_i\\
&= \sum\limits_{i=1}^{m}\left(r + \sum\limits_{j=1}^{m}M_{ij}x_j\right)p_i\\
&= r + \sum\limits_{i=1}^{m}p_i\left(\sum\limits_{j=1}^{m}M_{ij}x_j\right)
\end{align}
where we have used the fact that $\sum_i p_i = 1$ in the last line. Using \eqref{nD_replicator_mutator} to write down the equations for the frequencies $p_i$, we obtain
\begin{equation}
\frac{1}{N_K(t)}\frac{dp_i}{dt} = \left[(\mathbf{Mp})_i - \mathbf{p}\cdot\mathbf{Mp}\right]p_i   
\end{equation}
which is the familiar version of the replicator equation seen in most textbooks, with an extra $N_K(t)$ factor to account for the fact that $\sum_i x_i$ is allowed to fluctuate in our model. If instead $N_K$ was a constant for all time, it could simply be absorbed into the definition of the payoff matrix $M$ to obtain exactly the replicator equation as presented in most ecology/evolution textbooks. Both the stochastic dynamics \eqref{nD_example_SDE} and the deterministic limit \eqref{nD_example_det_limit} can be simplified from an $m$ dimensional system to an $m-1$ dimensional system by a coordinate transformation that projects the dynamics onto an appropriate curve: If we go from the variables $x_1,\ldots,x_m$ to the variables $p_1,\ldots,p_{m-1},N_K$, we can exploit the fact that $N_K$ varies much less than the $p_i$ terms to project the system onto a `slow manifold' in which $N_K$ is approximately constant, thus obtaining an $m-1$ dimensional system of equations and recovering the relation between the Lotka-Volterra equations for $m$ species and the replicator equation for $m-1$ tactics~\citep{constable_mapping_2017,parsons_dimension_2017}. However, I will not explore such dimensional reduction techniques further in this manuscript, and refer the reader to~\cite{constable_stochastic_2013} and~\cite{parsons_dimension_2017} for a review of the ideas of (stochastic) dynamics on slow manifolds.

>>>>

\section{An example in one dimension: The stochastic logistic equation}
Here, we analyze example \ref{ex_1D_stoch_logistic}. To recap, we had a population of individuals that exhibit a constant per-capita birth rate $\lambda > 0$, and a per-capita death rate that had the linear density-dependence $d(n) = \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n$, where $\mu$ and $K$ are positive constants. Thus, we have the equations
\begin{equation}
\label{ex_1D_stoch_logistic_BD_eqns}
\begin{aligned}
    b(n) &= \lambda n\\
    d(n) &= \left(\mu + (\lambda-\mu)\frac{n}{K}\right)n
\end{aligned}
\end{equation}
Here, $K$ is the system-size parameter. Introducing the new variable $x=n/K$, we obtain
\begin{align*}
    b_K(x) &= \frac{1}{K}b(n) = \frac{1}{K}\lambda Kx\\
    d_K(x) &= \frac{1}{K}d(n) = \frac{1}{K}\left(\mu + (\lambda-\mu)\frac{Kx}{K}\right)Kx
\end{align*}
Thus, we have
\begin{equation*}
    A^{\pm}(x) = b_K(x)\pm d_K(x) = x\left(\lambda \pm \left(\left(\mu + (\lambda-\mu)x\right)\right) \right)
\end{equation*}
Defining $r=\lambda-\mu$ and $v=\lambda+\mu$ and using equation \eqref{1D_SDE}, we 
see that the `mesoscopic view' of the system is given by the solution of the SDE
\begin{equation}\label{ex_1D_stoch_logistic_full_SDE}
dX_t =  rX_t(1-X_t)dt + \sqrt{\frac{X_t(v+rX_t)}{K}}dW_t
\end{equation}
From equation \eqref{1D_det_limit}, we see that the deterministic dynamics are
\begin{equation}\label{ex_1D_stoch_logistic_det_limit}
\frac{dx}{dt} = A^-(x) = rx(1-x)
\end{equation}
showing that in the infinite population limit, we obtain the logistic equation. This derivation also makes it clear that two systems with very different stochastic dynamics can nevertheless converge to the same infinite population limit \eqref{ex_1D_stoch_logistic_det_limit}, since \eqref{ex_1D_stoch_logistic_det_limit}t only depends on the difference $\lambda-\mu$.
\myfig{0.6}{Media/App_stoch_logistic_same_w_different_t.png}{Comparison of time series (Gillespie simulations) for two different parameter values of the stochastic logistic equation. Simulations with the parameter values $\lambda = 4, \mu = 2$ are plotted in red, and simulations with the parameter values $\lambda = 31, \mu = 29$ are plotted in blue. Both simulations have $K=200$. The infinite population limit of both simulations is plotted in black dotted lines. The graph shows 10 realizations each for the two sets of parameter values.}{fig_1D_same_w_different_t_comparison}
To illustrate the effects of this seemingly innocent fact, figure \ref{fig_1D_same_w_different_t_comparison} compares two simulations which have the same value of $\lambda - \mu$ but a ten-fold difference in $\lambda+\mu$. As is clear from the figure, though both populations have the same behavior at the infinite population limit, populations with a higher value of $\lambda+\mu$ exhibit much wilder fluctuations and are therefore more prone to stochastic extinction. This is the root cause of the noise-induced selection that occurs in higher dimensions, discussed in detail in part \ref{part_summary}.

@@

    dY_t &= A^-_1(t)Y_tdt + \sqrt{A^+_0(t)}dW_t\nonumber\\
    \Rightarrow dY_t &= r(1 - 2\alpha(t))Y_tdt + \sqrt{\alpha(t)(v+r\alpha(t))}dW_t\label{ex_1D_stoch_logistic_WNA}

@@

Let us imagine a population with $m$ types of individuals that are interacting according to some ecological rules. Let the state of the population be characterized by the vector $\mathbf{n}(t) = [n_1(t),n_2(t),\ldots,n_m(t)]$, where $n_i(t)$ is the number of type $i$ individuals at time $t$. Let the birth and death rates of the $i$th type be given by:
\begin{equation}
\label{nD_example_numbers_b_d_rates}
\begin{aligned}
b_i(\mathbf{n}) &= \left(\lambda + \frac{1}{K}\left(\sum\limits_{j=1}^{m}\beta_{ij}n_j\right)\right)n_i\\
d_i(\mathbf{n}) &= \left(\mu + \frac{1}{K}\left(\sum\limits_{j=1}^{m}\delta_{ij}n_j\right)\right)n_i
\end{aligned}
\end{equation}
where $K > 0$ is our system size parameter (and represents a global carrying capacity across all types), $\lambda > 0$ and $\mu > 0$ are suitable positive constants representing the baseline natality and mortality common to all types, and $\beta_{ij}$ and $\delta_{ij}$ are constants describing the effect of type $j$ individuals on the birth and death rate of type $i$ individuals respectively. The sign of $M_{ij} \coloneqq \beta_{ij} - \delta_{ij}$ determines whether type $j$ has a net positive or negative effect on the growth of type $i$. In ecological communities, this is a per-capita ecological interaction effect. In game-theoretic terms, we can interpret $M_{ij}$ as the payoff obtained by a type $j$ individual playing against a type $i$ individual. I assume that $| M_{ij} | \ll K$. The values $M_{ij}$ are often collected in an $m \times m$ matrix $\mathbf{M}$ called the `payoff matrix' (in evolutionary game theory) or `interaction matrix' (in Lotka-Volterra models). Lotka-Volterra models also frequently assume that the diagonal elements $M_{ii}$ are all equal, though I will not make that assumption here.

Going from population numbers $\mathbf{n}$ to densities $\mathbf{x} = \mathbf{n}/K$, we obtain the birth and death rates:
\begin{equation}
\label{nD_example_density_b_d_rates}
\begin{aligned}
b^{(K)}_i(\mathbf{x}) &= \left(\lambda + \sum\limits_{j=1}^{m}\beta_{ij}x_j\right)x_i\\
d^{(K)}_i(\mathbf{x}) &= \left(\mu + \sum\limits_{j=1}^{m}\delta_{ij}x_j\right)x_i
\end{aligned}
\end{equation}
Thus, we have
\begin{equation*}
A^{\pm}_{i} = x_i\left(\left(\lambda \pm \mu\right) + \sum\limits_{j=1}^{m}\left(\beta_{ij} \pm \delta_{ij}\right)x_j\right)
\end{equation*}
Defining $r \coloneqq \lambda - \mu$, $\nu \coloneqq \lambda + \mu$, and $T_{ij} \coloneqq \beta_{ij} + \delta_{ij}$, we see from equation \eqref{nD_Ito_SDE} that the mesoscopic view is the $m$ dimensional SDE given by
\begin{equation}
\label{nD_example_SDE}
d\mathbf{X}_{t} = \mathbf{A^-}(\mathbf{X}_t)dt + \frac{1}{\sqrt{K}}\mathbf{D}(\mathbf{X}_t)d\mathbf{W}_t
\end{equation}
where 
\begin{equation*}
\mathbf{A^-}_i = {(\mathbf{X}_{t})}_i(r + \sum\limits_{j=1}^{m}M_{ij}{(\mathbf{X}_{t})}_j) 
\end{equation*}
and
\begin{equation*}
(\mathbf{D}\mathbf{D}^{\mathrm{T}})_i = {(\mathbf{X}_{t})}_i(\nu + \sum\limits_{j=1}^{m}T_{ij}{(\mathbf{X}_{t})}_j) 
\end{equation*}
From \eqref{nD_det_limit}, we see that the deterministic limit is a set of $m$ coupled ODEs given by
\begin{equation}
\label{nD_example_det_limit}
\frac{d x_i}{dt} = x_i\left(r + \sum\limits_{j=1}^{m}M_{ij}x_j\right)
\end{equation}
These are precisely the Lotka-Volterra equations for a system of $m$ species. By matching the terms of \eqref{nD_example_density_b_d_rates} with those of \eqref{nD_functional_forms_for_replicator}, we can identify that we have $\mu = 0$ and
\begin{equation}
\label{nD_example_fitness}
w_i(\mathbf{x}) = r + \sum\limits_{j=1}^{m}M_{ij}x_j
\end{equation}
If $p_i(t)$ is the frequency of type $i$ individuals in the population at time $t$ and $N_K(t) = \sum_i x_i(t)$, then the mean fitness is given by
\begin{align}
\overline{w}(t) &= \sum\limits_{i=1}^{m}w_ip_i\\
&= \sum\limits_{i=1}^{m}\left(r + \sum\limits_{j=1}^{m}M_{ij}x_j\right)p_i\\
&= r + \sum\limits_{i=1}^{m}p_i\left(\sum\limits_{j=1}^{m}M_{ij}x_j\right)
\end{align}
where we have used the fact that $\sum_i p_i = 1$ in the last line. Using \eqref{nD_replicator_mutator} to write down the equations for the frequencies $p_i$, we obtain
\begin{equation}
\frac{1}{N_K(t)}\frac{dp_i}{dt} = \left[(\mathbf{Mp})_i - \mathbf{p}\cdot\mathbf{Mp}\right]p_i   
\end{equation}
which is the familiar version of the replicator equation seen in most textbooks, with an extra $N_K(t)$ factor to account for the fact that $\sum_i x_i$ is allowed to fluctuate in our model. If instead $N_K$ was a constant for all time, it could simply be absorbed into the definition of the payoff matrix $M$ to obtain exactly the replicator equation as presented in most ecology/evolution textbooks. Both the stochastic dynamics \eqref{nD_example_SDE} and the deterministic limit \eqref{nD_example_det_limit} can be simplified from an $m$ dimensional system to an $m-1$ dimensional system by a coordinate transformation that projects the dynamics onto an appropriate curve: If we go from the variables $x_1,\ldots,x_m$ to the variables $p_1,\ldots,p_{m-1},N_K$, we can exploit the fact that $N_K$ varies much less than the $p_i$ terms to project the system onto a `slow manifold' in which $N_K$ is approximately constant, thus obtaining an $m-1$ dimensional system of equations and recovering the relation between the Lotka-Volterra equations for $m$ species and the replicator equation for $m-1$ tactics~\citep{constable_mapping_2017,parsons_dimension_2017}. However, I will not explore such dimensional reduction techniques further in this manuscript, and refer the reader to~\cite{constable_stochastic_2013} and~\cite{parsons_dimension_2017} for a review of the ideas of (stochastic) dynamics on slow manifolds.
